---
title: "Stats 101A - Homework 10"
author: "Avelia Mira"
format:
  pdf:
    include-in-header:
      - text: |
          \usepackage{fvextra}
          \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: visual
---

## Load libraries & setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(car)
library(leaps)

births <- read_csv("birthsforHW.csv", show_col_types = FALSE)

# Convert Habit to factor if needed
if (!is.factor(births$Habit)) births$Habit <- as.factor(births$Habit)
```

## Homework 10

### 0) Fit the full model to predict baby birth weight

```{r}
m_full <- lm(weight ~ ., data = births)
summary(m_full)
```

- Full model: `weight ~ Gender + Premie + LowBirthWeight + Weeks + Apgar1 + Fage + Mage + Feduc + Meduc + TotPreg + Visits + Marital + Gained + Habit`
- Fit: $R^2 = 0.5276$, Adjusted $R^2 = 0.5233$.
- Strong predictors include `LowBirthWeightNot Low`, `Weeks`, `GenderMale`, `Gained`, `HabitSmoker`, `MaritalUnmarried`, and `TotPreg`.

### 1) Assess model validity (Residuals vs Fitted, Q-Q, Scale-Location)

```{r}
plot(m_full, which = 1:3)
```

- Based on diagnostics, the full model is not ideal: residual spread and Q-Q tail behavior suggest some non-constant variance and non-normality.
- This motivates trying a response transformation.

### 2) Confirm/refute trend assessment using marginal model plots

```{r}
marginalModelPlots(m_full, ~ Weeks + Apgar1 + Fage + Mage + Feduc + Meduc + TotPreg + Visits + Gained)
```

- Marginal model plots do not indicate severe nonlinearity in trend for the main numeric predictors.
- The larger issue is variance/normality rather than a strongly wrong mean trend form.

### 3) Check Box-Cox recommendation and usefulness of transforming weight

```{r}
boxCox(m_full)

# Optional alternative from lecture
inverseResponsePlot(m_full)

# Example transformed model (if Box-Cox suggests lambda near 0)
m_transformed <- lm(log(weight) ~ ., data = births)

par(mfrow = c(2, 2))
plot(m_transformed)
par(mfrow = c(1, 1))
```

- Using `log(weight)` improves model behavior; transformed-model fit is higher ($R^2 = 0.6612$, Adjusted $R^2 = 0.6581$) and diagnostics are more stable.
- I proceed with the transformed model as the more valid option.

### 4) Choose the most valid model and add Habit-specific intercept/slope with Gained

Use `Habit * Gained` to allow:
- different intercept by smoking habit
- different slope for `Gained` by smoking habit

```{r}
m_q4 <- lm(log(weight) ~ . + Habit:Gained, data = births)
summary(m_q4)
```

- Chosen working model for this part: `log(weight) ~ . + Habit:Gained`.
- Note: `Habit` main effect is already included via `.`, and `Habit:Gained` adds the interaction.

### 5) Best subsets on model (4), choose by BIC

```{r}
# Create model matrix to identify which columns to force in (Habit main effect)
X_full <- model.matrix(log(weight) ~ . + Habit:Gained, data = births)
habit_cols <- grep("^Habit", colnames(X_full))
habit_interaction_cols <- grep(":Habit|Habit:", colnames(X_full))
habit_main_cols <- setdiff(habit_cols, habit_interaction_cols)

# Force Habit main effect to stay in model (respecting hierarchy)
force_in_vec <- rep(FALSE, ncol(X_full) - 1)  # Exclude intercept
if (length(habit_main_cols) > 0) {
  force_in_vec[habit_main_cols - 1] <- TRUE  # Adjust for intercept
}

best_sub <- regsubsets(log(weight) ~ . + Habit:Gained, data = births, 
                       nvmax = 15, method = "exhaustive",
                       force.in = force_in_vec)
best_summary <- summary(best_sub)

min_bic_idx <- which.min(best_summary$bic)
print(paste("Best BIC is at index (number of predictors):", min_bic_idx))
best_summary$which[min_bic_idx, ]

plot(1:length(best_summary$bic), best_summary$bic, type = "b",
     xlab = "Model size", ylab = "BIC",
     main = "Best Subsets: BIC by Model Size")
```

- BIC-minimizing model index with forced Habit main effect.
- Final predictors include both `Habit` main effect and `Gained:Habit` interaction (respecting hierarchy).

### 6) Backward stepwise by BIC, compare with best subsets

```{r}
# Use same force.in vector to maintain hierarchy in backward selection
bwd_sub <- regsubsets(log(weight) ~ . + Habit:Gained, data = births, 
                      method = "backward", nvmax = 15,
                      force.in = force_in_vec)
bwd_summary <- summary(bwd_sub)

min_bwd_idx <- which.min(bwd_summary$bic)
bwd_summary$which[min_bwd_idx, ]
```

- Backward stepwise BIC with forced Habit main effect to respect hierarchy.
- Compare selected predictors with best subsets approach.

### 7) Fit the chosen best model and interpret effects

```{r}
# Fit the final model using the predictors selected by backward BIC in Q6
# Include Habit main effect to respect hierarchy principle
m_best <- lm(log(weight) ~ Gender + LowBirthWeight + Weeks + Apgar1 + Mage + TotPreg + Marital + Gained + Habit + Habit:Gained,
             data = births)

summary(m_best)
formula(m_best)
```

- Fitted final model:
  `log(weight) ~ Gender + LowBirthWeight + Weeks + Apgar1 + Mage + TotPreg + Marital + Gained + Habit + Habit:Gained`
- This model respects the hierarchy principle: both the main effect of `Habit` and the interaction `Habit:Gained` are included.
- 7a) `Mage` coefficient is positive, so (holding other variables fixed) each additional year of mother age is associated with a small increase in expected baby weight on the log scale.
- 7b) The `Habit` main effect provides a habit-specific intercept shift, while `Habit:Gained` allows the slope of weight gained to differ by smoking status—both intercept and slope vary by smoking habit.

```{r}
# Optional helper for smoking effect at average Gained
mean_gained <- mean(births$Gained, na.rm = TRUE)
coef_names <- names(coef(m_best))
coef_names[grepl("Habit", coef_names)]
mean_gained
```

### 8) Why are there 1910 fitted values instead of 2000 rows?

```{r}
nrow(births)
length(m_best$fitted.values)
nrow(births) - length(m_best$fitted.values)
```

`lm()` uses listwise deletion by default; rows with missing values in any used variable are removed.

For this fitted model: `nrow(births) = 2000`, `length(m_best$fitted.values) = 1539`, so `461` observations were removed due to missing data in at least one model variable.

### 9) First five rows removed by `lm`

```{r}
m_best$na.action
head(m_best$na.action, 5)
```

First five removed rows are `3, 9, 13, 14, 16`.

### 10) Identify bad leverage points

Use cutoff: `2*(p+1)/n`, where `n` is the number of observations actually used in the fitted model.

```{r}
p <- length(coef(m_best)) - 1
n_actual <- nobs(m_best)  # Use actual n after NA removal
cutoff <- 2 * (p + 1) / n_actual

leverages <- hatvalues(m_best)
std_res <- rstandard(m_best)
bad_idx <- which(leverages > cutoff & abs(std_res) > 2)

bad_leverage_table <- tibble(
  original_row = as.integer(names(leverages)[bad_idx]),
  row_after_na_omit = bad_idx,
  leverage = unname(leverages[bad_idx]),
  std_resid = unname(std_res[bad_idx])
)

print(paste("n after NA removal:", n_actual))
print(paste("p (excluding intercept):", p))
cutoff
bad_leverage_table
```

- Using `nobs(m_best)` to get the correct number of observations after listwise deletion.
- This ensures the leverage cutoff formula `2*(p+1)/n` uses the actual sample size of the fitted model.
- Bad leverage points are those with both high leverage (above cutoff) and large standardized residuals (|r| > 2).

### 11) View bad leverage rows correctly after NA removals

```{r}
births.na <- births[-m_best$na.action, ]

# Correct rows in data with omitted rows removed:
births.na[bad_idx, ]

# Optional quick summary of unusualness:
summary(births.na[bad_idx, ])
```

Explain why these observations might be unusual (extreme covariate values, combinations, or outcomes).

- These rows include unusually extreme birth profiles (for example, very low birth weights like `17`, `32`, `33` ounces and very short gestational ages such as `21–30` weeks), or uncommon combinations of predictors.
- Such extreme covariate patterns create high leverage, and when paired with large residuals they become bad leverage points.
