---
title: "Stats 101A - Take-Home Final Exam"
author: "Avelia Mira"
format: 
  pdf:
    pdf-engine: pdflatex
    include-in-header:
      - text: |
          \usepackage{fvextra}
          \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
          
editor: visual
---

# Q1 Variance Inflation and Model Fitting

This exam is based on a dataset comprised of characteristics of various LEGO sets. The data were scraped by Profs. Laura Ziegler and Anna Peterson at Iowa State from the website brickset.com. We will build a linear model that describes potential associations between the retail price and a set of predictors.

**Instructions:** Remove the variable Set_Name and fit the full model: price as a function of all available predictors. Do not carry out any transformations unless explicitly told to do so.

## Q1.1: Variance Inflation Factors

Identify three predictors in the full model that have "high" variance inflation factors.

**How to answer:** Look at the output from the `vif(m_full)` command. Find the three variables that have a VIF value greater than 5. List their names here.

**Answer:** Pieces, Unique_Pieces, weight

## Q1.2: P-value Change

The p-value for the t-test associated with Unique_Pieces has a large p-value (approx. 0.8). Suppose we refit the model but remove weight. How will the p-value for Unique_Pieces likely change?

**Options:** - The p-value will increase. - The p-value will decrease. - The p-value will stay the same.

**Answer:** **The p-value will decrease.**

## Q1.3: Explanation

Explain your answer to Q1.2 without refitting the model with weight removed.

**Answer:** Variance Inflation Factor (VIF) measures how severe the problem of collinearity is between predictor variables in a model. When variables are highly correlated (collinear), it inflates the variance, and therefore the standard error, of their estimated coefficients. Because the t-statistic is calculated by dividing the estimated slope by its standard error, an inflated standard error mathematically forces the t-statistic to become smaller. Removing the correlated variable `weight` reduces the collinearity for `Unique_Pieces`, which decreases its standard error, increases its t-statistic, and ultimately results in a smaller p-value.

## Q1.4: Code for Question 1

```{r q1-code}
# Load necessary libraries
library(tidyverse)
library(car)

# Load data, remove Set_Name, and drop missing values to ensure equal sample sizes
lego <- read_csv("lego.csv")
lego_clean <- lego |>
  select(-Set_Name) |> 
  drop_na()  # This ensures both models have the exact same rows!

# Q1.1 & Q2: Fit the full model and check VIF
m_full <- lm(price ~ ., data = lego_clean)
summary(m_full)
vif(m_full)

# Q2: Fit reduced model
m_reduced <- lm(price ~ Size + Pieces + Minifigures + Pages, data = lego_clean)
anova(m_reduced, m_full)
```

# Q2 Hypothesis Test for Joint Significance

Carry out the hypothesis test that weight and Unique_Pieces are jointly significant. That is, adding (or removing) both of them together has a statistically significant impact on the goodness of fit.

$$H_0: E(Y \mid X) = \beta_0 + \beta_1 \cdot \text{Size} + \beta_2 \cdot \text{Pieces} + \beta_3 \cdot \text{Minifigures} + \beta_4 \cdot \text{Pages}$$

$$H_a: E(Y \mid X) = \beta_0 + \beta_1 \cdot \text{Size} + \beta_2 \cdot \text{Pieces} + \beta_3 \cdot \text{Minifigures} + \beta_4 \cdot \text{Pages} + \beta_5 \cdot \text{Unique\_Pieces} + \beta_6 \cdot \text{weight}$$

## Q2.1: F-Statistic Value

What is the value of the F-Statistic for this test?

**How to answer:** Look at the `anova(m_reduced, m_full)` output from Q1.4. Look for the row corresponding to the full model and find the column labeled `F`. This is your F-statistic.

**Answer:** 238.89

## Q2.2: Statistical Decision

What decision do you make, using a significance level of $\alpha = 0.05$?

**Options:** - Reject the null and conclude that at least one of weight and Unique_Pieces are significant. - Fail to reject the null; there is insufficient evidence to conclude that at least one of weight and Unique_Pieces are significant.

**How to answer:** Look at the p-value column (`Pr(>F)`) in your `anova()` output. If the p-value is less than 0.05, choose the first option (reject the null). If the p-value is greater than or equal to 0.05, choose the second option (fail to reject the null).

**Answer:** Reject the null and conclude that at least one of weight and Unique_Pieces are significant.

## Q2.3: Code for Question 2

```{r q2-code}
# Fit the full model
m_full <- lm(price ~ ., data = lego_clean)

# Fit the reduced model (without weight and Unique_Pieces)
m_reduced <- lm(price ~ Size + Pieces + Minifigures + Pages, data = lego_clean)

# Conduct Partial F-test
anova(m_reduced, m_full)
```

The `anova(m_reduced, m_full)` command conducts a Partial F-test comparing the residual sums of squares between the two models.

# Q3 Best Model Selection

Assuming model validity, find the model that balances the trade-off between complexity and goodness of fit using BIC.

## Q3.1: Number of Variables

How many variables should there be in a model that minimizes the BIC?

**Answer:** 3

## Q3.2: Variables by Name

Identify these variables by name.

**Answer:** Size, Pages, weight

## Q3.3: Interpretation of Pages Coefficient

Fit the best model you identified in Q3.2 and write an interpretation of the slope for Pages. This should be no more than two, and ideally one, sentence long.

**Interpretation Template:** "Holding all other variables constant, for every 1-unit increase in Pages, the estimated mean price of the LEGO set changes by \[Insert Estimate\] dollars."

**Answer:** Holding all other variables constant, for every 1-unit increase in Pages, the estimated mean price of the LEGO set decreases by approximately 0.047 dollars.

## Q3.4: Code for Question 3

```{r q3-code}
library(leaps)

# Run best subsets on the full model
best_sub <- regsubsets(price ~ ., data = lego_clean, nvmax = 15, method = "exhaustive")
best_summary <- summary(best_sub)

# Find the model that minimizes BIC
min_bic_idx <- which.min(best_summary$bic)
print(paste("Number of variables minimizing BIC:", min_bic_idx))

# Identify the variables by name
best_summary$which[min_bic_idx, ]

# Plot BIC values
plot(1:length(best_summary$bic), best_summary$bic, type = "b",
     xlab = "Model size", ylab = "BIC",
     main = "Best Subsets: BIC by Model Size")

# Fit the best model using the variables identified above
# Based on Q3.2: Size, Pages, weight
m_best <- lm(price ~ Size + Pages + weight, data = lego_clean)
summary(m_best)
```

# Q4 Box-Cox Transformation

Investigate which transform of price would be optimal according to a Box-Cox Transform. Use the full model: `lm(price ~ ., data = lego_clean)`.

## Q4.1: Optimal Lambda Value

According to the Box-Cox transform, what is the optimal "rounded" response value for $\lambda$?

**How to answer:** Look at the output of `summary(powerTransform(m_full))`. Look under the column titled **`Rounded Pwr`**. That number is your optimal rounded response value for $\lambda$. (For example: if it is 0, the optimal transform is `log(price)`; if it is 0.5, it is `sqrt(price)`.)

**Answer:** 0.62

## Q4.2: Improvement in Trend Component

Does this transformation improve the validity of the trend component? Answer yes or no and explain.

**How to answer (mathematical check):** Use a formal functional-form test such as the Ramsey RESET test on both models. If the transformed model has a larger p-value (closer to non-significance), that is evidence of improved trend/linearity.

**Answer:** **No.** Using the Ramsey RESET test, the original model has $F=8.385$ with $p=2.73\times 10^{-4}$, while the transformed model has $F=133.481$ with $p=1.31\times 10^{-44}$. Both reject correct functional form, and the transformed model is more strongly rejected, so there is no mathematical evidence of trend-component improvement.

## Q4.3: Improvement in Normality Condition

Does this transformation improve the validity of the Normality condition? Answer yes or no and explain.

**How to answer (mathematical check):** Compare formal normality diagnostics (e.g., Shapiro-Wilk) and distribution shape metrics (skewness, kurtosis) for model residuals.

**Answer:** **Yes, partially.** Shapiro-Wilk improves from $W=0.7758$ ($p=1.11\times 10^{-22}$) in the original model to $W=0.8953$ ($p=1.17\times 10^{-15}$) in the transformed model. Residual skewness drops from $1.128$ to $0.622$, and kurtosis drops from $15.287$ to $8.576$. Normality is still rejected, but the transformed residuals are mathematically closer to normal.

## Q4.4: Improvement in Constant Variance Condition

Does this transformation improve the validity of the constant variance condition? Answer yes or no and explain.

**How to answer (mathematical check):** Compare a formal heteroskedasticity test (e.g., the Breusch-Pagan score test via `ncvTest`) between models.

**Answer:** **Yes, partially.** The score test statistic decreases from $\chi^2=689.69$ ($p=5.21\times 10^{-152}$) in the original model to $\chi^2=242.21$ ($p=1.30\times 10^{-54}$) in the transformed model. Heteroskedasticity is still statistically significant, but the severity is substantially reduced after transformation.

## Q4.5: Code for Question 4

```{r q4-code}
library(car)
library(lmtest)
library(tidyr)

# 1. Find the optimal Box-Cox transformation parameter
summary(powerTransform(m_full))

# 2. Fit the new transformed model 
# Based on lambda = 0.62, use price^0.62 transformation
m_trans <- lm(I(price^0.62) ~ ., data = lego_clean)

# 3. Mathematical diagnostics (instead of visual-only checks)

# 3a) Trend/functional form: Ramsey RESET
reset_full <- resettest(m_full, power = 2:3, type = "fitted")
reset_trans <- resettest(m_trans, power = 2:3, type = "fitted")
reset_full
reset_trans

# 3b) Normality: Shapiro-Wilk + shape metrics
sw_full <- shapiro.test(resid(m_full))
sw_trans <- shapiro.test(resid(m_trans))

skewness <- function(x) {
  m <- mean(x)
  s <- sd(x)
  mean((x - m)^3) / s^3
}

kurtosis <- function(x) {
  m <- mean(x)
  s <- sd(x)
  mean((x - m)^4) / s^4
}

normality_tbl <- data.frame(
  model = c("original", "transformed"),
  shapiro_W = c(unname(sw_full$statistic), unname(sw_trans$statistic)),
  shapiro_p = c(sw_full$p.value, sw_trans$p.value),
  skewness = c(skewness(resid(m_full)), skewness(resid(m_trans))),
  kurtosis = c(kurtosis(resid(m_full)), kurtosis(resid(m_trans)))
)
normality_tbl

# 3c) Constant variance: score test (Breusch-Pagan type)
ncv_full <- ncvTest(m_full)
ncv_trans <- ncvTest(m_trans)
ncv_full
ncv_trans

# 4. Optional visual diagnostics for supporting context
par(mfrow = c(2, 3))
plot(m_full, which = 1:3, main = "Original Model")
plot(m_trans, which = 1:3, main = "Transformed Model")
par(mfrow = c(1, 1))
```

# Q5 Assumptions

If you had to make any assumptions for any ambiguously worded questions, please explain here.

**Answer:** *(No assumptions needed; all questions are clearly specified.)*

# Q6 AI Tools Used

If you used AI, please state which AI models you used and where and how you used them. If none, write "none".

**Answer:** *(Fill in with your AI disclosure if applicable.)*