---
title: "Stats 102A - Homework 2 - Output File"
author: "Avelia Mira"
format: pdf
---

Copyright Miles Chen, Do not post, share, or distribute without permission.

To receive full credit the functions you write must pass all tests. We may conduct further tests that are not included on this page as well.

# Academic Integrity Statement

By including this statement, I, **Avelia Mira**, declare that all of the work in this assignment is my own original work.

At no time did I use an AI tool to generate or debug my solutions.

At no time did I look at the code of other students nor did I search for code solutions.

I understand that forbidden AI usage or plagiarism on any single part of this assignment will result in a 0 for the entire assignment.

\% Setting directory so that my code runs properly!

```{r, error = TRUE}
source("102A_hw_02_output_Avelia_Mira.R")  # edit with your file name
```

\pagebreak

# Part 1: Board representation

Create a single list object called `board` where you store the features of the game board in R.

```{r, error = TRUE}

# for this, I've decided to use a nested list approach. to be quite frank, I don't really care if we've learned this concept in class, but it's the first idea i came up with. so we're going with it

# okay nevermind -- this version of chutes and ladders doesn't have any overlapping chutes or ladders in one tile. but what i WAS going to do is basically just define characteristics of a tile and state if it has an endpoint or a start of a chute/ladder, and then denote it with an ID. so that way our code would be able to check if it's at the start of a chute or ladder, and if so, teleport to the end. similar to a JSON format

# our list shall have the following:
# rows, for the rows of the board
# cols, for the columns of the board
# ladders as a matrix. or basically a table, with one column being the start point and the other being the endpoint
# chutes as a matrix. technically same format as ladders, but i think it to be a good idea to denote them as separate parts of the list so that way it'll be easier to read code

board <- list(
  rows = 10,
  cols = 10,
  
  # ladders: start -> end (green arrows)
  ladders = matrix(c(
     1, 38,
     4, 14,
     9, 31,
    21, 42,
    28, 84,
    36, 44,
    51, 67, # haha. 67
    71, 91,
    80, 100
  ), ncol = 2, byrow = TRUE),
  
  # chutes: start -> end (red arrows)
  chutes = matrix(c(
    16,  6,
    47, 26, 
    49, 11,
    56, 53,
    62, 19,
    64, 60,
    87, 24,
    93, 73,
    95, 75,
    98, 78
  ), ncol = 2, byrow = TRUE)
)

# all of this should be verified. additionally, this makes it replicable and easy to use 
```

\pagebreak

# Part 2: Plot of Game board

```{r, error = TRUE}
# par() should help the plot be more visible. you can adjust this as necessary
par(mar = c(0, 0, 0, 0)) 
show_board(board)
```

\pagebreak

# Part 3: Miniboards

Create the `miniboard` objects and plots.

```{r, error = TRUE}
miniboard1 <- list(
  rows = 7,
  cols = 6,
  ladders = matrix(c(
    1, 23,
    9, 27,
    20, 31,
    33, 41
  ), ncol = 2, byrow = TRUE),
  chutes = matrix(c(
    16, 6,
    30, 18,
    35, 24,
    40, 26
  ), ncol = 2, byrow = TRUE)
)
par(mar = c(0, 0, 0, 0))
show_board(miniboard1)
```

```{r, error = TRUE}
miniboard2 <- list(
  rows = 9,
  cols = 7,
  ladders = matrix(c(
     9, 22,
    13, 30,
    24, 37,
    29, 41,
    33, 39,
    43, 54
  ), ncol = 2, byrow = TRUE),
  chutes = matrix(c(
    16, 3,
    31, 15,
    35, 21,
    62, 48
  ), ncol = 2, byrow = TRUE)
)
par(mar = c(0, 0, 0, 0))
show_board(miniboard2)
```

```{r, error = TRUE}
miniboard3 <- list(
  rows = 9,
  cols = 8,
  ladders = NULL,
  chutes = NULL
)
par(mar = c(0, 0, 0, 0))
show_board(miniboard3)
```

\pagebreak

# Part 4: Verbose output of one single player game

```{r, error = TRUE}
set.seed(5)
play_solo(board, verbose = TRUE)
```

\pagebreak

# Part 5: Monte Carlo Simulation Study

```{r, error = TRUE}
# run 10,000 games
n_simulations <- 10000


# initialise storage
vector_of_turns <- numeric(n_simulations)
ladder_9_usage <- logical(n_simulations)

# we need to accumulate the tallies for the barplots
# getting the dimensions from a single run to initialise
dummy_run <- play_solo(board, verbose = FALSE)
total_chute_counts <- numeric(length(dummy_run$chute_tally))
total_ladder_counts <- numeric(length(dummy_run$ladder_tally))

for (i in 1:n_simulations) {
  res <- play_solo(board, verbose = FALSE)
  
  # store turns
  vector_of_turns[i] <- res$turns
  
  # check if ladder 9 (the winning one) was used
  # it's the last ladder in our list, so index 9
  if (res$ladder_tally[9] > 0) {
    ladder_9_usage[i] <- TRUE
  }
  
  # add to total counts for barplots
  total_chute_counts <- total_chute_counts + res$chute_tally
  total_ladder_counts <- total_ladder_counts + res$ladder_tally
}

# create a custom summary with all the answers for the latter stuff
mc_stats <- list(
  min_turns     = min(vector_of_turns),
  count_min     = sum(vector_of_turns == min(vector_of_turns)),
  max_turns     = max(vector_of_turns),
  median_turns  = median(vector_of_turns),
  mean_turns    = mean(vector_of_turns),
  prop_over_100 = mean(vector_of_turns >= 100),
  prop_under_10 = mean(vector_of_turns <= 10),
  prop_ladder_9 = mean(ladder_9_usage) 
)
```

-   Create a histogram (breaks = 50) of the turns. Additionally, create barplots of chute and ladder tallies.

```{r}
hist(vector_of_turns, breaks = 50, 
     main = "Distribution of Game Lengths",
     xlab = "# of Turns")

barplot(total_chute_counts / n_simulations,
        names.arg = 1:length(total_chute_counts), # adds the 1, 2, 3 labels, taking the total_chute_counts as an argument
        main = "Relative Frequency of Chute Utilization", 
        xlab = "Chute ID", ylab = "Relative Frequency")

barplot(total_ladder_counts / n_simulations,
        names.arg = 1:length(total_ladder_counts), 
        main = "Relative Frequency of Ladder Utilization", 
        xlab = "Ladder ID", ylab = "Relative Frequency")
```

-   Find the minimum number of turns. How many times out of 10,000 did a game finish with the minimum number of turns?

```{r}
mc_stats$min_turns
mc_stats$count_min
```

-   Find the maximum number of turns.

```{r}
mc_stats$max_turns
```

-   What is the median number of turns?

```{r}
mc_stats$median_turns
```

-   What is the mean number of turns?

```{r}
mc_stats$mean_turns
```

-   What proportion of games take 100 or more turns to complete?

```{r}
mc_stats$prop_over_100
```

-   What proportion of games take 10 or fewer turns to complete?

```{r}
mc_stats$prop_under_10
```

-   What proportion of games utilize ladder 9 (the shortcut to win on space 80)?

```{r}
mc_stats$prop_ladder_9
```

# Optional Extra Credit

```{r}
probabiity_transition_matrix <- create_transition_matrix(board)

# calculate total states
total_states <- (board$rows * board$cols) + 1

state_0 <- numeric(total_states)
state_0[1] <- 1 

# compute the next states using our multiply_state() function
state_1 <- multiply_state(state_0, probabiity_transition_matrix)
state_2 <- multiply_state(state_1, probabiity_transition_matrix)
state_3 <- multiply_state(state_2, probabiity_transition_matrix)

# generate plots
par(mar = c(1, 1, 3, 1))
plot_prob_map(state_1, 1, board)
plot_prob_map(state_2, 2, board)
plot_prob_map(state_3, 3, board)
```