---
title: "Stats 101A - Homework 6"
author: "Avelia Mira"
format: 
  pdf:
    include-in-header:
      - text: |
          \usepackage{fvextra}
          \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}

editor: visual
---

## Load libraries & setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(broom)
setwd("~/Downloads/stats/stats101a/HW6")
```

## Question 1: Iowa Test Scores

### a) Iowa City schools vs. other schools

Create an indicator variable and fit a linear model to predict performance based on whether or not the school is in Iowa City. Interpret the summary output.

```{r}
iowa <- read_tsv("iowatest.txt")

iowa <- iowa |>
  mutate(IsIowaCity = if_else(City == "Iowa City", 1, 0))

# Fit linear model
model1a <- lm(Test ~ IsIowaCity, data = iowa)

# Summary
summary(model1a)
```

Schools in Iowa City score **14.7 points higher** on average (p = 0.000152). The R² = 0.104 shows location explains \~10% of score variation. This difference is statistically significant, indicating Iowa City schools perform better than other cities.

### b) Graphic comparing Iowa City schools to others

```{r}
# Create boxplot using ggplot
ggplot(iowa, aes(x = City, y = Test, fill = City)) +
  geom_boxplot() +
  labs(title = "Test Scores by City", x = "City", y = "Test Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

Iowa City shows a higher median test score than most other cities in the box plot.

### c) Poverty score and test scores

Fit a regression line to predict school test score from poverty score. Is there evidence that poverty is associated with the test score?

```{r}
# Fit model predicting test score from poverty score
model1c <- lm(Test ~ Poverty, data = iowa)

# Summary
summary(model1c)
```

Poverty is **strongly negatively associated** with test scores: each 1-unit increase in poverty decreases test score by 0.54 points (p \< 0.001). The R² = 0.673 shows poverty explains \~67% of score variation—much stronger than Iowa City. Schools in higher-poverty areas perform significantly worse.

### d) Diagnostic plots interpretation

Interpret the first three diagnostic plots: residuals plot, normal Q-Q plot, and scale-location plot.

```{r}
# Add diagnostics to the data
iowa_diag <- iowa |>
  mutate(
    fitted = fitted(model1c),
    residuals = residuals(model1c),
    std_resid = rstudent(model1c)
  )

# 1. Residuals vs Fitted
p1 <- ggplot(iowa_diag, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

# 2. Normal Q-Q Plot
p2 <- ggplot(iowa_diag, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()

# 3. Scale-Location Plot
model1c_aug <- augment(model1c)
p3 <- ggplot(model1c_aug, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Scale-Location", x = "Fitted Values", y = "Sqrt(|Std. Residuals|)") +
  theme_minimal()

# Display plots
p1
p2
p3
```

**Residuals vs Fitted:** Residuals scatter randomly around zero with uniform spread, suggesting linearity holds.

**Q-Q Plot:** Points follow the reference line closely (minor trailing at extremes), indicating residuals are approximately normal.

**Scale-Location:** The red line is roughly horizontal, suggesting constant variance across fitted values.

### e) Leverage points

Which point has the highest leverage? Are there any bad leverage points?

```{r}
# Calculate leverage using tidyverse
p <- length(coef(model1c))
n <- nrow(iowa)
threshold <- 2 * p / n

# Extract indices for sorting
leverage_vals <- hatvalues(model1c)
studentized_vals <- rstudent(model1c)

# Create leverage analysis dataframe
lev_analysis <- tibble(
  leverage = leverage_vals,
  studentized = studentized_vals,
  is_high = leverage > threshold,
  row_idx = seq_along(leverage)
) |>
  bind_cols(iowa) |>
  arrange(desc(leverage))

# Display highest leverage point
cat("Threshold for high leverage (2p/n):", round(threshold, 4), "\n\n")
cat("Highest leverage point:\n")
print(lev_analysis |> slice(1) |> select(School, City, Poverty, leverage, studentized))

# High leverage points
cat("\nAll high leverage points:\n")
print(lev_analysis |> filter(is_high) |> select(School, City, Poverty, leverage))
```

The highest leverage point is Longfel (Waterloo) with poverty = 99 (most extreme X value). However, **no bad leverage points** exist because Longfel's residual is reasonable and fits the trend well—high leverage alone isn't problematic.

### f) F-test hypotheses and conclusion

State the null and alternative hypotheses for the F-test in the summary output from Question (c). Give the conclusion at 5% significance level.

The null hypothesis, Ho, is β₁ = 0, that poverty score has no effect on test scores.

The alternative hypothesis, Ha, is β₁ ≠ 0, that poverty score DOES affect test scores.

The F-statistic is 269.79, the p-value is 2.2e-16, essentially 0, and we use a significance level of α = 0.05. As such, we reject the null hypothesis at the 5% significance level. We believe there is evidence that that poverty score is significantly associated with test scores.

------------------------------------------------------------------------

## Question 2: Housing Indicators

Load the `indicators.txt` file and fit a model to predict PriceChange from LoanPaymentsOverdue.

```{r}
#| message: false
# Load and parse indicators data (requires careful parsing due to file structure)
indicators_raw <- readLines("indicators.txt")
indicators_raw <- gsub('\r', '', indicators_raw)
lines <- strsplit(indicators_raw, "\t")
data_list <- lapply(lines[-1], function(x) x[x != ""])
metro_areas <- sapply(data_list, function(x) x[1])
price_change <- as.numeric(sapply(data_list, function(x) x[2]))
loan_payments <- as.numeric(sapply(data_list, function(x) x[3]))
indicators <- data.frame(PriceChange = price_change, LoanPaymentsOverdue = loan_payments)
rownames(indicators) <- metro_areas

# Fit model
model2 <- lm(PriceChange ~ LoanPaymentsOverdue, data = indicators)
summary(model2)
```

Each 1% increase in loan payments is associated with a 2.25% **decrease** in price change (p = 0.024). The R² = 0.279 shows loan payments explain \~28% of price variation. The model is statistically significant.

### a) Trend and constant variance (Residuals vs Fitted plot)

Using the first plot from `plot.lm()`, comment on whether the trend component is fit correctly and whether the constant variance condition appears to hold.

```{r}
# Residuals vs Fitted plot using ggplot
model2_diag <- augment(model2)

ggplot(model2_diag, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

The residuals show a **curved U-pattern** (negative → positive → negative), indicating the linear trend is misspecified. Variance also increases at higher fitted values (heteroscedasticity). Both the trend and constant variance assumptions **fail**.

### b) Q-Q plot comparison with simulated normal data

Generate 19 Q-Q norm plots from a N(0,1) distribution. Compare the Q-Q norm plot from the data to this collection.

```{r}
# Q-Q plot using ggplot
ggplot(model2_diag, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()

# Generate simulated Q-Q plots for comparison
set.seed(123)
n_sims <- nrow(indicators)

# Create a tibble of simulated data
simulated_data <- tibble(
  sim_id = rep(1:19, each = n_sims),
  residuals = c(replicate(19, rnorm(n_sims, mean = 0, sd = 1)))
)

# Plot simulated Q-Q plots
ggplot(simulated_data, aes(sample = residuals)) +
  stat_qq(size = 2) +
  stat_qq_line(color = "red") +
  facet_wrap(~sim_id, ncol = 5) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7))
```

Simulated Q-Q plots follow the reference line closely (normal), while the actual data shows a **systematic S-curve**, deviating significantly in the tails. The errors are **not normally distributed**.

### c) Validity - Normal distribution of errors

Comment on the validity of the model with respect to the condition that the errors are Normally distributed.

The S-shaped Q-Q plot indicates residuals are **not normally distributed**. This violates a key linear regression assumption and makes hypothesis tests and confidence intervals unreliable.

### d) Scale-Location plot - constant variance

Using the Scale-Location plot, comment on model validity with respect to the constant variance condition.

```{r}
# Scale-Location plot using ggplot
ggplot(model2_diag, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Scale-Location", x = "Fitted Values", y = "Sqrt(|Standardized Residuals|)") +
  theme_minimal()
```

The red line follows a U-shaped pattern—variance is lower at mid-range and higher at extremes. This **non-constant variance (heteroscedasticity)** violates the constant variance assumption, making standard errors unreliable and invalidating hypothesis tests.

### e) High leverage points

Identify any high leverage points (which row are they in). State the criteria you are using.

```{r}
# Calculate leverage using tidyverse
p2 <- length(coef(model2))
n2 <- nrow(indicators)
threshold2 <- 2 * p2 / n2

# Create leverage analysis
lev_analysis2 <- model2_diag |>
  mutate(
    leverage = .hat,
    studentized = rstudent(model2),
    is_high = leverage > threshold2,
    metro_area = rownames(indicators)
  ) |>
  arrange(desc(leverage)) |>
  select(metro_area, leverage, PriceChange, LoanPaymentsOverdue, studentized, is_high)

cat("Threshold for high leverage (2p/n):", round(threshold2, 4), "\n\n")
cat("High leverage points:\n")
print(lev_analysis2 |> filter(is_high) |> select(metro_area, leverage, PriceChange, LoanPaymentsOverdue))

cat("\nTop 3 leverage points:\n")
print(lev_analysis2 |> slice(1:3) |> select(metro_area, leverage))
```

Using threshold 2p/n = 0.222, high leverage points are: **MiamiFt.Lauderdale** (0.285, highest loan payments) and **Seattle** (0.240, lowest loan payments).

### f) Bad leverage points

Identify any bad leverage points (if there are any). If there are none, say so.

```{r}
# Check for bad leverage points using filter
bad_leverage2 <- lev_analysis2 |>
  filter(is_high == TRUE & abs(studentized) > 2)

if (nrow(bad_leverage2) > 0) {
  cat("Bad leverage points found:\n")
  print(bad_leverage2 |> select(metro_area, leverage, studentized))
} else {
  cat("NO bad leverage points identified.\n")
  cat("All high leverage points have reasonable residuals (|residual| < 2).\n")
}
```

Although MiamiFt.Lauderdale and Seattle have high leverage, **no bad leverage points** exist because their residuals fit the trend well.

### g) High influence points (Cook's Distance)

Identify any points with high influence, as measured by Cook's Distance. Explain why you think these are high.

```{r}
# Calculate Cook's Distance using tidyverse
threshold_cook <- 4 / n2

# Create influence analysis
inf_analysis <- model2_diag |>
  mutate(
    cooks_d = cooks.distance(model2),
    is_high = cooks_d > threshold_cook,
    metro_area = rownames(indicators)
  ) |>
  arrange(desc(cooks_d)) |>
  select(metro_area, cooks_d, is_high, everything())

cat("Threshold for high influence (4/n):", round(threshold_cook, 4), "\n\n")
cat("High influence points (Cook's D >", round(threshold_cook, 4), "):\n")
print(inf_analysis |> filter(is_high) |> select(metro_area, cooks_d))

cat("\nTop 3 most influential points:\n")
print(inf_analysis |> slice(1:3) |> select(metro_area, cooks_d))
```

Using threshold 4/n = 0.222, **Seattle** is high influence (Cook's D = 0.495). It has extreme X value (1.65% loan payments) and large residual (6.9% price change vs \~0.8% fitted), making it highly influential. Removing Seattle would substantially change model parameters.

### h) Influence vs. Leverage - In your own words

What do influence and leverage measure?

**Leverage** measures how extreme a point's X value is—its *potential* to affect the model. **Influence** (Cook's Distance) measures the *actual* impact—combining both leverage and residual size. High leverage doesn't guarantee influence; influence requires both extreme X *and* poor fit. Seattle exemplifies this: high leverage + large residual = high influence.

------------------------------------------------------------------------

## Question 3: Airfare Data Analysis (Chapter 3, Problem 1)

The data file airfares.txt on the book web site gives the one-way airfare (in US dollars) and distance (in miles) from city A to 17 other cities in the US. Interest centers on modeling airfare as a function of distance. The first model fit to the data was

Given below and in Figure 3.41 is some output from fitting model (3.7).

### R Output from Model (3.7)

```         
Call:
lm(formula = Fare ~ Distance)

Coefficients:
                Estimate    Std. Error    t value    Pr(>|t|)
(Intercept)     48.971770   4.405493      11.12      1.22e-08 ***
Distance        0.219687    0.004421      49.69      <2e-16 ***

Signif. codes: 0'***'0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 10.41 on 15 degrees of freedom
Multiple R-Squared: 0.994, Adjusted R-squared: 0.9936
F-statistic: 2469 on 1 and 15 DF, p-value: < 2.2e-16
```

**Model Equation:** $$\widehat{\text{Fare}} = 48.97 + 0.220 \times \text{Distance}$$

**Interpretation of Coefficients:** - **Intercept (48.97):** Base fare (intercept) is approximately \$48.97. - **Distance coefficient (0.220):** Each additional mile increases fare by approximately \$0.22. - **t-statistic (49.69):** Extremely large, indicating Distance is highly significant. - **p-value (\< 2e-16):** Essentially zero — Distance coefficient is highly statistically significant.

### (a) Critique of the analyst’s conclusion

The analyst concluded: “The regression coefficient of the predictor variable, Distance is highly statistically significant and the model explains 99.4% of the variability in the Y-variable, Fare. Thus model (1) is a highly effective model for both understanding the effects of Distance on Fare and for predicting future values of Fare given the value of the predictor variable, Distance.”

This conclusion is **overly optimistic**. Statistical significance and a high $R^2$ do not guarantee that the model form is correct or that predictions are reliable. The diagnostic plot (Figure 3.41) shows a systematic curved pattern in the residuals versus Distance, indicating **nonlinearity** and **model misspecification**. Because the residuals are not randomly scattered around zero, the straight-line model produces **systematic bias** (underpredicting short and long distances, overpredicting mid-range distances). As a result, the model is **not** “highly effective” for understanding the true relationship or for prediction across the range of distances. Good inference requires checking assumptions, not just $R^2$ and p-values.

### (b) Fit of the straight-line model and how to improve it

No, the straight-line model **fails**—residuals show clear curvature. Add a **quadratic term** or use **log transformation** to capture the nonlinear relationship. These would yield randomly scattered residuals and better predictions.